Active-contours driven registration method for structure-informed segmentation of diffusion MR images

### 2015-04-15 Publications/2015-NeuroImage/00-main.tex
change kw
updates to export to text
added Gert's comments
updates and standalone figures
regseg label
add regseg label uppercase
add Meri's comments

\documentclass[5p,authoryear]{elsarticle}
% \documentclass[authoryear,onecolumn]{elsarticle}
\journal{ArXiv archiving for pre-publication review}

% *** MISC UTILITY PACKAGES ***
\usepackage{glossaries}
\usepackage{booktabs} % to use \toprule
\usepackage{multirow}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{xifthen}
\usepackage{xspace}

%\usepackage[cmex10,intlimits]{amsmath}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{algorithmic}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\argmax}{argmax}
\DeclareMathOperator{\argmin}{argmin}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\dist}{dist}
\DeclareMathOperator{\const}{Const.}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}}
\providecommand{\mdist}[2]{ \mathcal{D}_{#2}^2(\mathbf{#1}) }
\providecommand{\omegaset}{\ensuremath{\boldsymbol{\Omega}}}
\providecommand{\gammaset}{\ensuremath{\boldsymbol{\Gamma}}}

@@

\usepackage[T1]{fontenc}
\usepackage{charter}
%\usepackage[expert]{mathdesign}
%\usepackage[libertine,cmintegrals,cmbraces,vvarbb]{newtxmath}
%\renewcommand*{\familydefault}{\sfdefault}
\usepackage[switch]{lineno}
\usepackage{array}
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

\address[bit]{Biomedical Image Technologies (BIT), ETSI Telecomunicaci\'on, %
Universidad Polit\'ecnica de Madrid, Madrid, Spain}
\address[ciber]{Centro de Investigaci\'on Biom\'edica en Red en Bioingenier\'ia, Biomateriales y Nanomedicina (CIBER-BBN), Zaragoza, Spain}
\address[ucla]{Department of Mathematics, University of California,
Los Angeles (UCLA), Los Angeles, CA, US}
\address[scil]{Computer Science Department, Faculty of Science, Universit\'e de Sherbrooke, 2500 Boulevard Universit\'e, Sherbrooke, QC J1K 2R1, Canada}
\address[lts5]{Signal Processing Laboratory (LTS5), \'Ecole Polytechnique
F\'ed\'erale de Lausanne (EPFL), Lausanne, Switzerland}
\address[chuv]{Dept. of Radiology, University
Hospital Center (CHUV) and University of Lausanne (UNIL), Lausanne, Switzerland}

@@

% Note that keywords are not normally used for peerreview papers.
\begin{keyword}
diffusion MRI \sep cortical~parcellation \sep segmentation %
\sep nonlinear~registration \sep active~contours \sep echo-planar imaging.
\end{keyword}

>>>>

% \documentclass[5p,authoryear,fleqn]{elsarticle}
\documentclass[3p,authoryear,fleqn]{elsarticle}
\journal{ArXiv archiving for pre-publication review}

% *** MISC UTILITY PACKAGES ***
\usepackage[mode=buildnew]{standalone}
\usepackage{gincltex}
\usepackage{glossaries}
\usepackage{booktabs} % to use \toprule
\usepackage{multirow}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{xifthen}
\usepackage{xspace}
\usepackage{graphicx}
\graphicspath{{figures/}}

%\usepackage[cmex10,intlimits]{amsmath}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{algorithmic}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\argmax}{argmax}
\DeclareMathOperator{\argmin}{argmin}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\dist}{dist}
\DeclareMathOperator{\const}{Const.}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}}
\providecommand{\mdist}[2]{ \mathcal{D}_{#2}^2(\mathbf{#1}) }
\providecommand{\omegaset}{\ensuremath{\boldsymbol{\Omega}}}
\providecommand{\gammaset}{\ensuremath{\boldsymbol{\Gamma}}}
\providecommand{\regseg}{\emph{regseg}}
\providecommand{\Regseg}{\emph{Regseg}}

@@

\usepackage[T1]{fontenc}
\usepackage{charter}
%\usepackage[expert]{mathdesign}
%\usepackage[libertine,cmintegrals,cmbraces,vvarbb]{newtxmath}
%\renewcommand*{\familydefault}{\sfdefault}

\usepackage{array}
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

\usepackage{ifpdf}

\ifpdf
  \usepackage[switch]{lineno}
  % nothing to do
\else
  \makeatletter
  \def\temp{dvips.def}
  \ifx\Gin@driver\temp
  \def\Ginclude@graphics#1{\def\temp{#1}---image \expandafter\strip@prefix\meaning\temp---}
  \fi
  \makeatother
\fi

@@

\address[bit]{Biomedical Image Technologies (BIT), ETSI Telecomunicaci\'on, %
Universidad Polit\'ecnica de Madrid, Madrid, Spain}
\address[ciber]{Centro de Investigaci\'on Biom\'edica en Red en Bioingenier\'ia, Biomateriales y Nanomedicina (CIBER-BBN), Zaragoza, Spain}
\address[ucla]{Department of Mathematics, University of California,
Los Angeles (UCLA), Los Angeles, CA, US}
\address[scil]{Computer Science Department, Faculty of Science, Universit\'e de Sherbrooke, 2500 Boulevard Universit\'e, Sherbrooke, QC J1K 2R1, Canada}
\address[lts5]{Signal Processing Laboratory (LTS5), \'Ecole Polytechnique
F\'ed\'erale de Lausanne (EPFL), Lausanne, Switzerland}
\address[chuv]{Dept. of Radiology, CIBM, University
Hospital Center (CHUV) and University of Lausanne (UNIL), Lausanne, Switzerland}

@@

% Note that keywords are not normally used for peerreview papers.
\begin{keyword}
diffusion MRI \sep cortical~parcellation \sep segmentation %
\sep nonlinear~registration \sep active~contours \sep susceptibility~distortion.
\end{keyword}

### 2015-04-15 Publications/2015-NeuroImage/03-introduction.tex
add Meri's comments

% -*- root: 00-main.tex -*-
\section{Introduction}\label{sec:introduction}
The accurate delineation of \gls*{wm} in \gls*{dmri} and the fusion of prior
  anatomical information extracted from a \gls*{t1} image of the same subject
  are crucial in a range of applications based on tractography, such as
  the extraction of structural connectivity \citep{craddock_imaging_2013} or
  tract-based spatial statistics \citep{smith_tractbased_2006}.
However, segmenting \gls*{dmri} data precisely, and coregistration of anatomical
  images are difficult due to several limitations.
First, \gls{dmri} images suffer from partial voluming of several tissues in 
  voxels located at their interfaces due to the low resolution of images
  (generally around \isores[mm]{2.2}).
Second, \gls*{dmri} schemes probe the diffusion process within the brain in 
  many angles called \glspl*{dwi}, completed by one or more baseline (\emph{b0}) 
  volumes without directional gradients.
The extremely low \gls*{snr} and high dimensionality of \glspl*{dwi} disable their
  direct use in segmentation.
The low contrast between \gls*{gm} and \gls*{wm} in the \emph{b0} volumes is not suitable for 
  segmentation either.
Lastly, \gls*{dmri} images are acquired using \gls*{epi} to speed up acquisition
  at the cost of introducing geometrical distortion and signal degradation and
  destruction \citep{jezzard_correction_1995}.
The artifact has been proven to cause an important impact on the anatomy extracted
  from \gls*{dmri}, particularly in certain fiber bundles \citep{irfanoglu_effects_2012}.
These limitations preclude segmentation in native \gls*{dmri} space, including
  registration-based approaches using anatomical images as source of structural
  information.

Early attempts to delineate the \gls*{wm} proceeded by thresholding the 
  \gls*{fa}\footnote{the \gls*{fa} is a scalar parameter of diffusion derived from
  the \gls*{dmri} data.} map.
However, the mask and subsequent analyses substantially depend on the threshold set
  \citep{taoka_fractional_2009} owing to the large variability of \gls*{fa} and its
  decay with the increase of incoherence between fibers.
\cite{zhukov_level_2003} proposed active contours with edges represented
  by levelsets, evolving on directionally invariant scalar maps.
\cite{rousson_level_2004} successfully segmented the corpus callosum with
  region-based levelsets on the eigenvalues of tensors fitted on the
  \gls*{dmri} data.
A large body of work targeted the definition of appropriate features to cluster,
  such as the 5D representation proposed by \cite{jonasson_segmentation_2005}.
Other efforts include mixture models on sets of directionally invariant maps
  \citep{liu_brain_2007}, iterative \citep{hadjiprocopis_unbiased_2005} and
  hierarchical \citep{lu_segmentation_2008} clusterings,
  graph-cuts \citep{han_experimental_2009},
  and volume fraction modeling \citep{kumazawa_improvement_2013}.

@@

Therefore, given the described circumstances, a method performing simultaneously
  segmentation in native \gls*{dmri} space and registration of the corresponding \gls*{t1} image
  to the \gls*{dmri} data could provide the optimal solution.
Previously, joint segmentation and registration has been successfully applied to
  longitudinal object tracking \citep{paragios_level_2003} and atlas-based
  segmentation \citep{gorthi_active_2011}.
The most extended approach is integrating active contours in image registration 
  methods.
\cite{unal_coupled_2005} and later \cite{wang_joint_2006},
  improved an existing method \citep{yezzi_variational_2003} based on linear registration
  to the nonlinear case implementing a free-form deformation field.
\cite{droske_mumfordshah_2009} reviewed the existing techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration and segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
Recently, \cite{guyader_combined_2011} proposed a simultaneous segmentation and
  registration method in 2D using levelsets and a nonlinear elasticity smoother on the
  displacement vector field, which preserves topology even with very large deformations.
Finally, \cite{gorthi_active_2011} extended the existing methodologies using a multiphase
  levelset function for the registration of several active contours, in the application
  of atlas-based segmentation.

>>>>

% -*- root: 00-main.tex -*-
\section{Introduction}\label{sec:introduction}
% Current applications to investigate the fiber pathways of the brain
%   using \gls*{dmri}, such as tract-based spatial statistics \citep{smith_tractbased_2006}
%   and connectome extraction \citep{craddock_imaging_2013}, critically rely on the fusion of 
%   prior anatomical information extracted from a \gls*{t1} image of the same subject.
The accurate delineation of \gls*{wm} in \gls*{dmri} and the fusion of prior
  anatomical information extracted from a \gls*{t1} image of the same subject
  are crucial in a range of applications based on tractography, such as
  the extraction of structural connectivity \citep{craddock_imaging_2013} or
  tract-based spatial statistics \citep{smith_tractbased_2006}.
However, segmenting \gls*{dmri} data precisely, and coregistration of anatomical
  images are difficult due to several limitations.
First, \gls{dmri} images present a resolution much higher than the imaged
  microstructural features \citep{basser_microstructural_1996}.
Therefore, voxels located at structural discontinuities are affected by partial
  voluming of sources of signal.
Second, \gls*{dmri} schemes probe the diffusion process within the brain in 
  many angles called \glspl*{dwi}, completed by one or more baseline (\emph{b0}) 
  volumes without directional gradients.
The extremely low \gls*{snr} and high dimensionality of \glspl*{dwi} disable their
  direct use in segmentation.
The low contrast between \gls*{gm} and \gls*{wm} in the \emph{b0} volumes is not suitable for 
  segmentation either.
Lastly, \gls*{dmri} images are acquired using \gls*{epi} to speed up acquisition
  at the cost of introducing geometrical distortion and signal degradation and
  destruction \citep{jezzard_correction_1995}.
The artifact has been proven to cause an important impact on the anatomy extracted
  from \gls*{dmri}, particularly in certain fiber bundles \citep{irfanoglu_effects_2012}.
These limitations preclude segmentation in native \gls*{dmri} space, including
  registration-based approaches using anatomical images as source of structural
  information.

Early attempts to delineate the \gls*{wm} proceeded by thresholding the 
  \gls*{fa}\footnote{the \gls*{fa} is a scalar parameter of diffusion derived from
  the \gls*{dmri} data.} map.
However, the mask and subsequent analyses substantially depend on the threshold set
  \citep{taoka_fractional_2009}.
To overcome the unreliability of of \gls*{fa} thresholding,
  \cite{zhukov_level_2003} proposed active contours with edges represented
  by levelsets, evolving on directionally invariant scalar maps.
\cite{rousson_level_2004} successfully segmented the corpus callosum with
  region-based levelsets on the eigenvalues of tensors fitted on the
  \gls*{dmri} data.
A large body of work targeted the definition of appropriate features to cluster,
  such as the 5D representation proposed by \cite{jonasson_segmentation_2005}.
Other efforts include mixture models on sets of directionally invariant maps
  \citep{liu_brain_2007}, iterative \citep{hadjiprocopis_unbiased_2005} and
  hierarchical \citep{lu_segmentation_2008} clusterings,
  graph-cuts \citep{han_experimental_2009},
  and volume fraction modeling \citep{kumazawa_improvement_2013}.

@@

Therefore, given the described circumstances, a method performing simultaneously
  segmentation in native \gls*{dmri} space and registration of the corresponding \gls*{t1} image
  to the \gls*{dmri} data could provide the optimal solution.
As far as we know, such strategy has never been proposed for the application at hand.
Previously, joint segmentation and registration has been successfully applied to other problems.
For instance, longitudinal object tracking \citep{paragios_level_2003} and atlas-based
  segmentation \citep{gorthi_active_2011}.
The most extended approach is integrating active contours in image registration 
  methods.
\cite{unal_coupled_2005} and later \cite{wang_joint_2006},
  improved an existing method \citep{yezzi_variational_2003} based on linear registration
  to the nonlinear case implementing a free-form deformation field.
\cite{droske_mumfordshah_2009} reviewed the existing techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration and segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
Recently, \cite{guyader_combined_2011} proposed a simultaneous segmentation and
  registration method in 2D using levelsets and a nonlinear elasticity smoother on the
  displacement vector field, which preserves topology even with very large deformations.
Finally, \cite{gorthi_active_2011} extended the existing methodologies using a multiphase
  levelset function for the registration of several active contours, in the application
  of atlas-based segmentation.
